---
title: Fatal Shootings by U.S. Police Since 2015
author: Isabella Benabaye
date: '2020-06-27'
categories: [rstats, data visualization, reactable]
tags: [police killings, rstats, data visualization, the washington post, reactable]
summary: 'Looking at the number of fatal shootings in the U.S. by on-duty police officers in every state and cities with at least 10 deaths. Data from [The Washington Post](https://github.com/washingtonpost/data-police-shootings).'
reading_time: yes
external_link: ''
image:
  caption: ''
  focal_point: ''
  preview_only: true
toc: true
links:
 - name: "code"
   url: https://github.com/isabellabenabaye/isabella-b.com/blob/master/content/blog/us-fatal-shootings-police/index.Rmd
   icon_pack: fab
   icon: github
 - name: "source"
   url: https://github.com/washingtonpost/data-police-shootings
output: hugodown::hugo_document
---

The other day, I came across [The Washington Post's database](https://github.com/washingtonpost/data-police-shootings) of every **fatal shooting** in the United States by a **police officer in the line of duty** since January 1, 2015. 

About the people and cases recorded, from their [`readme`](https://github.com/washingtonpost/data-police-shootings/blob/master/README.md):
<blockquote>The Post is documenting only those shootings in which a police officer, in the line of duty, shot and killed a civilian — the circumstances that most closely parallel the 2014 killing of Michael Brown in Ferguson, Mo., which began the protest movement culminating in Black Lives Matter and an increased focus on police accountability nationwide. The Post is not tracking deaths of people in police custody, fatal shootings by off-duty officers or non-shooting deaths.</blockquote>

I pulled the data and created the interactive tables below, where you can search for a state or city and see the number of killings in that area. By default, they are sorted by the total number of deaths per place, but you can also sort them by the number of deaths by race.

<lastupdate> Tables last updated: `r Sys.time()` </lastupdate>

### Fatal shootings in every <span style="color:#3fc1c9">U.S. state</span> since 2015
```{r message=FALSE, warning=FALSE, include=FALSE}
library(tidyverse)
library(reactable)
library(htmltools)

fatal_police_shootings <- read_csv("https://raw.githubusercontent.com/washingtonpost/data-police-shootings/master/fatal-police-shootings-data.csv")

# bar chart
bar_chart <- function(label, width = "100%", height = "18px", fill = "#3fc1c9", background = NULL) {
  bar <- div(style = list(background = fill, width = width, height = height))
  chart <- div(style = list(flexGrow = 1, marginLeft = "6px", background = background), bar)
  div(style = list(display = "flex", alignItems = "center"), label, chart)
}

# state counts pre-pivot - used for cell shading
state_counts <- fatal_police_shootings %>%
  count(state, race, sort = TRUE)

# state name data - get full state names
states <- bind_cols(datasets::state.name, datasets::state.abb) %>%
  rename(state_name = "...1",
         state = "...2")

# get percents
state_data <- fatal_police_shootings %>%
  left_join(states) %>%
  count(state_name, race) %>%
  ## make a column for each race
  pivot_wider(id_cols = state_name, names_from = race, values_from = n, values_fill = 0) %>%
  rename(unknown = `NA`) %>%
  mutate(total = A + B + N + W + unknown + H + O,
         W_perc = W/total *100,
         B_perc = B/total *100,
         H_perc = H/total *100,
         A_perc = A/total *100,
         N_perc = N/total *100,
         O_perc = O/total *100,
         unknown_perc = unknown/total *100)

# main table
state_table <- state_data %>%
  select(state_name, total, W, H, B, A, N, O, unknown)

# palette
red_pal <- function(x) rgb(colorRamp(c("#ffeef1", "#f87274"), bias = 2)(x), maxColorValue = 255)

# race column
race_column <- function(col, class = NULL, ...){
  colDef(
    ...,
    maxWidth = 80,
    defaultSortOrder = "desc",
    headerStyle = list(fontSize = "13px", fontWeight = 400),
    style = function(value) {
      normalized <- (value - min(state_counts[-1:-2,"n"])) / (max(state_counts[-1:-2,"n"]) - min(state_counts[-1:-2,"n"]))
      if (value < 1) {
        list(color = "#aaa", background = "#F7F7F7")
      } else if (normalized > 0.99) {
        list(background = "#f87274")
      } else {
        list(background = red_pal(normalized))
      }
    }
  )
}

# grouped columns
race_cols <- c("W", "B", "H", "A", "N", "O", "unknown")


# reactable table
final_table_states <- reactable(
  data = state_table,
  style = list(fontFamily = "Work Sans, sans-serif", fontSize = "14px", background = "#F7F7F7"),
  defaultSorted = "total",
  defaultPageSize = 15,
  defaultColDef = colDef(class = "cell", headerClass = "header"),
  highlight = TRUE,
  searchable = TRUE,
  compact = TRUE,
  columnGroups = list(
    colGroup(name = "Race of person killed", columns = race_cols)
  ),
  columns = list(
    state_name = colDef(name = "State",
                        minWidth = 140,
                        style = list(fontWeight = 500)),
    total = colDef(
      name = "Number of killings",
      defaultSortOrder = "desc",
      minWidth = 160,
      cell = function(value) {
        width <- paste0(value * 100 / max(state_table$total), "%")
        value <- format(value, big.mark = ",")
        # Fix each label using the width of the widest number (incl. thousands separators)
        value <- format(value, width = 4, justify = "right")
        bar_chart(value, width = width)
      },
      # And left-align the columns
      align = "left",
      style = list(fontFamily = "DM Mono", whiteSpace = "pre")
    ),
    W = race_column(name = "White", col = "W"),
    B = race_column(name = "Black", col = "B"),
    H = race_column(name = "Hispanic", col = "H"),
    A = race_column(name = "Asian", col = "A"),
    N = race_column(name = "Native American", col = "N"),
    O = race_column(name = "Other", col = "O"),
    unknown = race_column(name = "Unknown", col = "unknown")
  )
)
```

`r final_table_states`
<br>
**California** has a much higher number of fatal shootings as compared to Texas, with over 300 more deaths. **Hispanics** are highly disproportionally affected in both states, as well as in Arizona and New Mexico. 

{{% alert look %}}
It's important to note that these tables show the **raw counts** and don't include the populations of the areas, which you need to consider to have a holistic view of these cases. [This The Washington Post article](https://www.washingtonpost.com/graphics/investigations/police-shootings-database/) takes that into account and reports that New Mexico, Alaska, and Oklahoma have the highest rates of shootings. 
{{% /alert %}} 

To narrow down the 2,798 cities in the database, I've only included the cities with 10 or more killings.

### <span style="color:#FCA311">U.S. cities</span> with at least 10 fatal shootings
```{r message=FALSE, warning=FALSE, include=FALSE}
# Cities table -----
# city counts pre-pivot - used for cell shading
city_counts <- fatal_police_shootings %>%
  mutate(city = paste0(city,", ",state)) %>%
  count(city, race, sort = TRUE)

# main table
city_table <- fatal_police_shootings %>%
  mutate(city = paste0(city,", ",state)) %>%
  count(city, race) %>%
  ## make a column for each race
  pivot_wider(id_cols = city, names_from = race, values_from = n, values_fill = 0) %>%
  rename(unknown = `NA`) %>%
  mutate(total = A + B + N + W + unknown + H + O) %>%
  select(city, total, W, H, B, A, N, O, unknown) %>%
  filter(total >= 10)

# race column
race_column <- function(col, class = NULL, ...){
  colDef(
    ...,
    maxWidth = 80,
    defaultSortOrder = "desc",
    headerStyle = list(fontSize = "13px", fontWeight = 400),
    style = function(value) {
      normalized <- (value - min(city_counts[-1:-2,"n"])) / (max(city_counts[-1:-2,"n"]) - min(city_counts[-1:-2,"n"]))
      if (value < 1) {
        list(color = "#aaa", background = "#F7F7F7")
      } else if (normalized > 0.99) {
        list(background = "#f87274")
      } else {
        list(background = red_pal(normalized))
      }
    }
  )
}

# reactable table
final_table_cities <- reactable(
  data = city_table,
  style = list(fontFamily = "Work Sans, sans-serif", fontSize = "14px", background = "#F7F7F7"),
  defaultSorted = "total",
  defaultPageSize = 15,
  defaultColDef = colDef(class = "cell", headerClass = "header"),
  highlight = TRUE,
  searchable = TRUE,
  compact = TRUE,
  columnGroups = list(
    colGroup(name = "Race of person killed", columns = race_cols)
  ),
  columns = list(
    city = colDef(name = "State",
                  minWidth = 140,
                  style = list(fontWeight = 500)),
    total = colDef(
      name = "Number of killings",
      defaultSortOrder = "desc",
      minWidth = 160,
      cell = function(value) {
        width <- paste0(value * 100 / max(city_table$total), "%")
        value <- format(value, big.mark = ",")
        # Fix each label using the width of the widest number (incl. thousands separators)
        value <- format(value, width = 4, justify = "right")
        bar_chart(value, width = width, fill = "#FCA311")
      },
      # And left-align the columns
      align = "left",
      style = list(fontFamily = "DM Mono", whiteSpace = "pre")
    ),
    W = race_column(name = "White", col = "W"),
    B = race_column(name = "Black", col = "B"),
    H = race_column(name = "Hispanic", col = "H"),
    A = race_column(name = "Asian", col = "A"),
    N = race_column(name = "Native American", col = "N"),
    O = race_column(name = "Other", col = "O"),
    unknown = race_column(name = "Unknown", col = "unknown")
  )
)
```

`r final_table_cities`
<br>
**Los Angeles**, the city with the most deaths in the state with the most deaths, has almost four times more killings than Bakersfield, the city with the second-most in California. To see the cities in a certain state, for example, you can search: `, CA`.


### Racial demographics of the deaths vs. the <span style="color:#0EAD69">U.S. Population</span>
```{r message=FALSE, warning=FALSE, include=FALSE}
population <- tribble(
  ~race, ~perc, ~race_full,
  "W", 60.1, "White",
  "H", 18.5, "Hispanic",
  "B", 13.4, "Black",
  "A", 5.9, "Asian",
  "N", 1.3, "Native American",
  "O",0.8, "Other"
)

# pop data + overall deaths demographics
population <- fatal_police_shootings %>%
  count(race) %>%
  summarise(race, perc_killed = n/sum(n) *100) %>%
  inner_join(population) %>%
  arrange(desc(perc)) %>%
  select(race_full, perc, perc_killed)

pop_table <- reactable(population,
  style = list(fontFamily = "Work Sans, sans-serif", fontSize = "15px", background = "#F7F7F7"),
  columns = list(
    race_full = colDef(
      name = "",
      width = 150,
      style = list(fontWeight = 500)
    ),
    perc = colDef(
      name = "% of the population",
      defaultSortOrder = "desc",
      minWidth = 200,
      cell = function(value) {
        width <- paste0(value * 100 / max(population$perc), "%")
        value <- paste0(value, "%")
        # Fix each label using the width of the widest number (incl. thousands separators)
        value <- format(value, width = 5, justify = "right")
        bar_chart(value, width = width, fill = "#0EAD69")
      },
      # And left-align the columns
      align = "left",
      style = list(fontFamily = "DM Mono", whiteSpace = "pre"),
      format = colFormat(percent = TRUE)
    ),
    perc_killed = colDef(
      name = "% of those killed",
      defaultSortOrder = "desc",
      minWidth = 200,
      cell = function(value) {
        width <- paste0(value * 100 / max(population$perc), "%")
        value <- paste0(round(value), "%")
        # Fix each label using the width of the widest number (incl. thousands separators)
        value <- format(value, width = 5, justify = "right")
        bar_chart(value, width = width, fill = "#525252")
      },
      # And left-align the columns
      align = "left",
      style = list(fontFamily = "DM Mono", whiteSpace = "pre"),
      format = colFormat(percent = TRUE)
    )
  )
)
```
`r pop_table`
<br>
Overall, the Black community appears to clearly be disproportionately affected by this kind of violence. To see where they have been particularly affected, you can sort the number of killings per city by the number of Black people killed.

The Washington Post [article](https://www.washingtonpost.com/graphics/investigations/police-shootings-database/) I mentioned earlier presents their findings from the data, including a great plot that visualizes how Black Americans are killed at a rate twice as high as that of white Americans. They also discuss how the rate of shootings has remained steady each year and map where they occur and the number of shootings per million people.



### More about the data
<blockquote>In 2015, The Post began tracking more than a dozen details about each killing — including the race of the deceased, the circumstances of the shooting, whether the person was armed and whether the victim was experiencing a mental-health crisis — by culling local news reports, law enforcement websites and social media and by monitoring independent databases such as Killed by Police and Fatal Encounters. The Post conducted additional reporting in many cases.</blockquote>

Read more about The Washington Post's project and check out the data at their GitHub repository [<i class="fas fa-external-link-square-alt"></i>](https://github.com/washingtonpost/data-police-shootings)

